---
title: "Reading Q-View project files with comparsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading Q-View project files with comparsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(comparsR)
```

`comparsR` is a pure-R parser for `.Q-View` binary project files. The
file format is a single-file container that bundles an embedded H2 SQL
database with binary plate-image data; the rendered CSV report is
stored as a CLOB inside the database. `read_qview()` extracts that
report without needing a Java runtime or H2 driver.

## Reading a project file

```{r, eval = FALSE}
qv <- read_qview("path/to/project.Q-View")
qv
```

The returned object has class `qview` and contains:

```{r, eval = FALSE}
str(qv, max.level = 1)
#> List of 11
#>  $ metadata          :List of 12
#>  $ manifest          : tibble [4 x 3]
#>  $ segments          : tibble [3 x 4]
#>  $ analytes          : tibble [<n_analytes> x 8]
#>  $ well_groups       : tibble [<n_groups> x 7]
#>  $ pixel_intensities : tibble [<n_rows> x 8]
#>  $ summary_statistics: tibble [<n_rows> x 6]
#>  $ concentrations    : NULL or tibble
#>  $ curve_fit         : NULL or tibble
#>  $ report_csv        : character vector
#>  $ plate_layout      : tibble [96 x 7]
#>  - attr(*, "class") = "qview"
```

## Output tibbles

The shape of each table (with placeholder analyte names):

### `analytes`

```
spot_number | analyte    | unit  | lod | lloq | uloq | assay_control_low | assay_control_high
-----------------------------------------------------------------------------------------
1           | analyte_1  | ng/ml | NA  | NA   | NA   | NA                | NA
2           | analyte_2  | ug/ml | NA  | NA   | NA   | NA                | NA
...
```

### `well_groups`

```
well_group | sample_id | is_standard | is_negative | is_sample | is_control | well_type
----------------------------------------------------------------------------------------
Cal 1      | Cal 1     | TRUE        | FALSE       | FALSE     | FALSE      | standard
Low        | Low       | FALSE       | TRUE        | FALSE     | FALSE      | negative
sample_a   | sample_a  | FALSE       | FALSE       | TRUE      | FALSE      | sample
...
```

### `pixel_intensities`

```
well_group | sample_id | well | replicate | analyte   | unit  | pixel_intensity | dilution
-------------------------------------------------------------------------------------------
sample_a   | sample_a  | C3   | 1         | analyte_1 | ng/ml | 12345           | 100
sample_a   | sample_a  | C3   | 1         | analyte_2 | ug/ml |  4567           | 100
...
```

## The naming convention

When Q-View imports a plate-template CSV it rewrites the identifiers:

* `Cal 1` -> `ICal 1`
* `Low`   -> `GLow`
* `High`  -> `HHigh`
* `FD...` and any all-digit ID -> a single `N` prefix.

Pass `strip_prefix = TRUE` to undo the rewrite:

```{r, eval = FALSE}
qv <- read_qview("path/to/project.Q-View", strip_prefix = TRUE)
qv$well_groups$sample_id
```

## Plotting

```{r, eval = FALSE}
plot(qv, type = "plate_map")
plot(qv, type = "intensity_heatmap")
plot(qv, type = "replicate_scatter")
```

`plot.qview()` requires `ggplot2`.

## Exporting

```{r, eval = FALSE}
qview_to_xlsx(qv, "out.xlsx")
qview_to_csv_dir(qv, "out_csv/")
saveRDS(qv, "out.rds")
```

## Interactive front-end

```{r, eval = FALSE}
qview_app()
```

Launches a small Shiny app: upload a `.Q-View` file (and optionally a
plate-template CSV), preview each parsed table, and download the
result as `xlsx`, `rds`, or a zip of CSVs.
