---
title: "Reading and analysing .Q-View files with qviewparsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and analysing .Q-View files with qviewparsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
library(qviewparsR)
```

## What this package does

`qviewparsR` is a pure-R parser for the binary `.Q-View` project file
format used in chemiluminescent multiplex ELISA plate imaging and
quantification. A single `.Q-View` file bundles:

* a plain-text manifest header,
* an embedded H2 SQL database (with project metadata, the analyte
  panel, plate geometry, sample-to-well assignments, replicate pixel
  intensities, and -- if the user generated a report inside the
  producing application -- a fully rendered CSV report stored as a
  CLOB), and
* one or more binary LOB segments holding the raw chemiluminescent
  plate images.

`read_qview()` extracts everything except the raw images and returns
it as a list of tidy tibbles, with no Java runtime, no H2 database
driver, and no compiled code anywhere in the package.

## Installation

```{r}
# install.packages("pak")
pak::pak("CTTIR/qviewparsR")
```

`qviewparsR` requires R >= 4.1.0 plus a small set of tidyverse-aligned
dependencies (`cli`, `dplyr`, `lifecycle`, `openxlsx2`, `readr`,
`rlang`, `tibble`, `tidyr`). Plotting requires `ggplot2` (Suggested);
the Shiny front-end additionally requires `shiny`, `bslib`, and `DT`.

## A complete walk-through

The end-to-end workflow is short:

```{r}
library(qviewparsR)

qv <- read_qview("path/to/plate.Q-View")
qv                          # one-screen summary

qv$analytes                 # spot_number, analyte, unit, lod, lloq, uloq
qv$well_groups              # one row per sample/calibrator/control
qv$pixel_intensities        # long-format replicate readings
qv$summary_statistics       # per-group mean / std-dev / CV rows
qv$plate_layout             # one row per plate well

summary(qv)                 # mean / SD / CV per well type x analyte
```

`read_qview()` always returns a list of class `qview` with eleven slots
described in `?read_qview`. Empty slots are zero-row tibbles rather
than `NULL`, so downstream code can rely on shape stability.

## The naming convention

The producing software rewrites identifiers from the original
well-assignment template CSV before it stores them. The mapping is
systematic and reversible:

| Template value                | Stored as          |
|-------------------------------|--------------------|
| `Cal 1` ... `Cal N`           | `ICal 1` ... `ICal N` |
| `Low`                         | `GLow`             |
| `High`                        | `HHigh`            |
| `FD24277364`, `1211498458`, ...| `NFD24277364`, `N1211498458`, ... |

`strip_qview_prefix()` reverses the rewrite. Pass `strip_prefix = TRUE`
to `read_qview()` to apply it across every sample-id column at once:

```{r}
qv <- read_qview("path/to/plate.Q-View", strip_prefix = TRUE)
unique(qv$well_groups$sample_id)
```

The vectorised helper is also useful on its own:

```{r}
strip_qview_prefix(c("ICal 1", "GLow", "HHigh", "NFD24277364"))
#> [1] "Cal 1"      "Low"        "High"       "FD24277364"
```

## Coercion and tidy-data idioms

`as_tibble()` returns the long-format `pixel_intensities` table -- the
primary tabular payload -- so a parsed object can drop straight into a
dplyr / ggplot2 pipeline:

```{r}
library(dplyr)
library(tibble)

qv |>
  as_tibble() |>
  filter(replicate == 1L) |>
  group_by(analyte, unit) |>
  summarise(median_pi = median(pixel_intensity, na.rm = TRUE),
            .groups = "drop")
```

The `is_qview()` predicate lets package-aware functions guard their
inputs:

```{r}
is_qview(qv)      # TRUE
is_qview(list())  # FALSE
```

## Visualisation

Three quick-look plot types are built in (ggplot2 required):

```{r}
plot(qv, type = "plate_map")          # 96-well plate, fill = well type
plot(qv, type = "intensity_heatmap")  # facet per analyte, fill = PI
plot(qv, type = "replicate_scatter")  # rep 1 vs rep 2 per analyte
```

Each call returns a `ggplot` object, so themes, scales, and labels can
be added on top:

```{r}
library(ggplot2)
plot(qv, type = "plate_map") +
  theme_bw(base_size = 12) +
  labs(title = NULL, subtitle = "QC overview")
```

## Exporting

Three writers cover the common destinations. All return the parsed
object invisibly, so they compose with `|>`:

```{r}
qv |>
  write_qview_xlsx("plate.xlsx") |>      # one sheet per parsed table
  write_qview_csv ("plate_csv/") |>      # one CSV file per parsed table
  write_qview_rds ("plate.rds")          # full lossless R round-trip
```

Add `overwrite = TRUE` to `write_qview_xlsx()` / `write_qview_rds()` to
replace existing destinations.

The legacy aliases `qview_to_xlsx()` and `qview_to_csv_dir()` still
work but are flagged with `lifecycle::deprecate_warn()` and will be
removed in a future release.

## Cross-validating against a template CSV

Every well assignment is already embedded in the Q-View file. If you
also have the original well-assignment template CSV the producing
application imported, `read_qview_template()` parses it into a tibble
that aligns with `qv$plate_layout` for cross-validation:

```{r}
tmpl <- read_qview_template("path/to/template.csv")

qv$plate_layout |>
  dplyr::left_join(tmpl, by = "well", suffix = c("_qview", "_template")) |>
  dplyr::filter(sample_id_qview != sample_id_template)
```

Any rows surviving the filter expose template-vs-Q-View mismatches.

## Interactive front-end

For non-coding collaborators, `qview_app()` launches a small Shiny
application with the same workflow exposed visually:

```{r}
qview_app()
```

The app accepts a `.Q-View` upload (and optionally a template CSV),
shows every parsed table and the three plots in tabs, and offers
one-click downloads as `xlsx`, `rds`, or a zipped CSV directory.

## Error handling

Every exported function validates its inputs early and raises a
structured `cli::cli_abort()` error that points at the user's call,
not at internal helpers. Typical shapes you may see:

```
Error in `read_qview()`:
! `path` must be an existing file.
x "missing.Q-View" does not exist.
```

```
Error in `read_qview()`:
! `path` is not a valid `.Q-View` project file.
x "junk.bin" is missing the expected container header.
i Expected a numeric container version followed by "Q-View Project".
```

The messages carry an `i` bullet whenever there is an actionable hint.

## Where to go next

* `?read_qview` -- exhaustive description of every output slot.
* `?write_qview` -- the three exporters share one help page.
* `?summary.qview` -- detail on the per-well-type aggregation.
* `?qview_app` -- launching the interactive app.

The `inst/extdata/` directory does **not** ship a Q-View fixture
because the binary format is large; instead, the test suite skips
cleanly when no fixture is present (`tests/testthat/test-read_qview.R`
documents the lookup paths).
